Reanalysis Step 2: Multiple Imputation

### 2025-08-19 reanalysis-2-multiple_imputation.qmd
wip

```{r setup}
#| include: false
# Setup analysis environment (includes seed setting)
source("dependencies.R")
setup_analysis(seed = 2025)

@@

# Set parameters based on mode
if (analysis_mode == "test") {
  n_imputations <- 2
  n_iterations <- 5
  message("Running in TEST mode: ", n_imputations, " imputations, ", n_iterations, " iterations")
} else if (analysis_mode == "final") {
  n_imputations <- 30
  n_iterations <- 20
  message("Running in FINAL mode: ", n_imputations, " imputations, ", n_iterations, " iterations")
} else {
  stop("Invalid analysis_mode. Must be 'test' or 'final'")
}
```

@@

# Extract variable lists
ps_model_vars <- variable_lists$ps_model_vars
outcome_model_vars <- variable_lists$outcome_model_vars
```

@@

```{r define-imputation-vars}
# Variables to include in imputation model (following Rubin's guidance)
# Include: outcome, exposure, all covariates SOURCE variables only
# Derived variables will be recreated post-imputation

# Core analysis variables - SOURCE ONLY
imputation_vars <- c(
  # Patient ID (not imputed, just kept)
  "e_mrn_deidentified",

@@

  # Demographics/Clinical - SOURCE ONLY
  "baseline_VA_logMAR",
  "gender_cat",
  "race_ethnic_cat",
  "insurance_cat",
  "age",  # Continuous age (source)
  # NOT age_cat - derived from cut(age)
  "CCI",
  "DCSI",

@@

# Select only imputation variables
df_mice <- df_for_imputation |>
  dplyr::select(dplyr::all_of(imputation_vars)) |>
  mutate(
    # Convert person_dr to ordered factor for polr imputation method
    person_dr = factor(person_dr, levels = c(0, 1, 2), 
                      labels = c("No_DR", "NPDR", "PDR"),
                      ordered = TRUE)
  )
```

## Assess Missing Data Pattern

```{r missing-pattern}
# Visualize missing data pattern
missing_pattern <- md.pattern(df_mice, rotate.names = TRUE)

# Summary of missingness
miss_summary <- df_mice |>
  miss_var_summary() |>
  filter(n_miss > 0)

miss_summary |>
  kable(caption = "Variables with missing data")

# Visualize missingness relationships
gg_miss_upset(df_mice, nsets = 10)
```

@@

# Print initial imputation model summary
message("=== IMPUTATION MODEL SETUP ===\n")
message("Total variables in dataset: ", ncol(df_mice))
message("Variables to be imputed: ", sum(mice_dry$nmis > 0))
message("Total missing values: ", sum(mice_dry$nmis))
message("\nMissing data by variable:")
miss_vars <- data.frame(
  Variable = names(mice_dry$nmis),
  N_Missing = mice_dry$nmis,
  Pct_Missing = round(mice_dry$nmis/nrow(df_mice) * 100, 1)
) |>
  filter(N_Missing > 0) |>
  arrange(desc(N_Missing))

@@

# Check default imputation methods
method_table <- data.frame(
  Variable = names(mice_dry$method),
  Method = mice_dry$method,
  stringsAsFactors = FALSE
) |>
  filter(Method != "")

@@

# Use predictive mean matching for continuous/count variables
custom_methods["CCI"] <- "pmm"  # Predictive mean matching for count data
custom_methods["DCSI"] <- "pmm"  # Predictive mean matching for count data
custom_methods["baseline_VA_logMAR"] <- "pmm"
custom_methods["age"] <- "pmm"
custom_methods["outcome_VA_logMAR"] <- "pmm"  # If any missing

@@

# Don't impute these variables (but they ARE still used as predictors!)
# Method = "" means: use as predictor but don't impute
custom_methods["e_mrn_deidentified"] <- ""  # ID - never impute
custom_methods["ever_lapse_binary"] <- ""  # Complete - use as predictor only
custom_methods["glaucoma_bef_hitplus_cat"] <- ""  # Already recoded to 0/1
custom_methods["otherretina_bef_hitplus_cat"] <- ""  # Already recoded to 0/1
custom_methods["gender_cat"] <- ""  # Likely complete
custom_methods["other_inject"] <- ""  # Likely complete
custom_methods["anti_VEGF"] <- ""  # Likely complete  
custom_methods["focal_laser_flag"] <- ""  # Likely complete
custom_methods["PRP_flag"] <- ""  # Likely complete

# Note: Variables with method = "" are STILL USED as predictors for other variables
# They just aren't imputed themselves

@@

# Don't use ID as predictor
pred_matrix[, "e_mrn_deidentified"] <- 0
pred_matrix["e_mrn_deidentified", ] <- 0

>>>>

```{r setup}
#| include: false
source("dependencies.R")
setup_analysis(seed = 2025)

@@

# Set parameters based on mode
if (analysis_mode == "test") {
  n_imputations <- 2
  n_iterations <- 5
  message("Running in TEST mode: ", n_imputations, " imputations, ", n_iterations, " iterations")
} else if (analysis_mode == "final") {
  n_imputations <- 10
  n_iterations <- 20
  message("Running in FINAL mode: ", n_imputations, " imputations, ", n_iterations, " iterations")
} else {
  stop("Invalid analysis_mode. Must be 'test' or 'final'")
}
```

@@

ps_model_vars <- variable_lists$ps_model_vars
outcome_model_vars <- variable_lists$outcome_model_vars
```

@@

```{r define-imputation-vars}
# Variables to include in imputation model
# Include: outcome, exposure, all covariates source variables only
# Derived variables will be recreated post-imputation

# Core analysis variables - SOURCE ONLY
imputation_vars <- c(
  "e_mrn_deidentified",

@@

  # Demographics/Clinical
  "baseline_VA_logMAR",
  "gender_cat",
  "race_ethnic_cat",
  "insurance_cat",
  "age",  # Continuous age (source)
  # NOT age_cat - derived from cut(age)
  "CCI",
  "DCSI",

@@

df_mice <- df_for_imputation |>
  dplyr::select(dplyr::all_of(imputation_vars)) |>
  mutate(
    # Convert person_dr to ordered factor for polr imputation method
    person_dr = factor(person_dr, levels = c(0, 1, 2), 
                      labels = c("No_DR", "NPDR", "PDR"),
                      ordered = TRUE)
  )
```

@@

# Print initial imputation model summary
message("Total variables in dataset: ", ncol(df_mice))
message("Variables to be imputed: ", sum(mice_dry$nmis > 0))
message("Total missing values: ", sum(mice_dry$nmis))
message("\nMissing data by variable:")
miss_vars <- data.frame(
  Variable = names(mice_dry$nmis),
  N_Missing = mice_dry$nmis,
  Pct_Missing = round(mice_dry$nmis/nrow(df_mice) * 100, 1)
) |>
  filter(N_Missing > 0) |>
  arrange(desc(N_Missing))

@@

method_table <- data.frame(
  Variable = names(mice_dry$method),
  Method = mice_dry$method,
  stringsAsFactors = FALSE
) |>
  filter(Method != "")

@@

custom_methods["CCI"] <- "pmm"  # Predictive mean matching for count data
custom_methods["DCSI"] <- "pmm"  # Predictive mean matching for count data
custom_methods["baseline_VA_logMAR"] <- "pmm"
custom_methods["age"] <- "pmm"
custom_methods["outcome_VA_logMAR"] <- "pmm"

@@

# Don't impute these variables
# Method = "" means: don't impute (but still use as predictor EXCEPT for ID)
custom_methods["e_mrn_deidentified"] <- ""  # ID - never impute, never use as predictor
custom_methods["ever_lapse_binary"] <- ""  # Complete - use as predictor only
custom_methods["glaucoma_bef_hitplus_cat"] <- ""  # Already recoded to 0/1
custom_methods["otherretina_bef_hitplus_cat"] <- ""  # Already recoded to 0/1
custom_methods["gender_cat"] <- ""  # Likely complete
custom_methods["other_inject"] <- ""  # Likely complete
custom_methods["anti_VEGF"] <- ""  # Likely complete  
custom_methods["focal_laser_flag"] <- ""  # Likely complete
custom_methods["PRP_flag"] <- ""  # Likely complete

# Note: Variables with method = "" (except ID) are STILL USED as predictors for other variables
# They just aren't imputed themselves. The ID is excluded from both imputation and prediction.

@@

# Don't use ID as predictor AND don't impute it
pred_matrix[, "e_mrn_deidentified"] <- 0
pred_matrix["e_mrn_deidentified", ] <- 0
